---
title: "Paid vs. Organic Search for the CPP Farm Store"
subtitle: "AO4 Proposal Presentation: Website SEO and Google Shopping Ads"
author: "Julieta Diaz"
institute: "IBM 6500, Cal Poly Pomona"
date: "`r Sys.Date()`"
format:
  revealjs:
    theme: [serif, farmstore.scss]
    css: styles.css
    width: 1600
    height: 900
    logo: images/logo.svg
    footer: "CPP Farm Store CEP | AO4: Paid vs. Organic Search"
    transition: slide
    transition-speed: default
    background-transition: fade
    incremental: false
    toc: true
    toc-depth: 1
    toc-title: "Agenda"
    slide-number: true
    menu: true
    preview-links: auto
    scrollable: true
    smaller: true
    code-fold: false
    code-overflow: wrap
    number-sections: false
    number-depth: 5
    embed-resources: true
editor: visual
bibliography: references.bib
csl: american-marketing-association.csl
execute:
  echo: true
  warning: false
  message: false
  freeze: auto
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(scales)
library(knitr)
```

```{r}
#| label: get-data
#| include: false
# ------------------------------------------------------------------
# DATA SOURCE: Google's public GA4 sample e-commerce dataset
# (Google Merchandise Store, Nov 1 2020 to Jan 31 2021) in BigQuery.
#
# The query only runs when the cached CSV does not exist yet.
# Run this chunk ONCE interactively in RStudio so bigrquery can open
# the browser and log you in. After that, rendering just reads the CSV.
# ------------------------------------------------------------------
data_file <- "data/ga4_channel_funnel.csv"

if (!file.exists(data_file)) {
  library(bigrquery)
  billing_project <- "ibm6500-bigquery"   # Google Cloud project ID used to run the query

  sql <- "
    SELECT
      traffic_source.medium AS medium,
      COUNT(DISTINCT user_pseudo_id) AS users,
      COUNT(DISTINCT IF(event_name = 'view_item',      user_pseudo_id, NULL)) AS viewed_item,
      COUNT(DISTINCT IF(event_name = 'add_to_cart',    user_pseudo_id, NULL)) AS added_to_cart,
      COUNT(DISTINCT IF(event_name = 'begin_checkout', user_pseudo_id, NULL)) AS began_checkout,
      COUNT(DISTINCT IF(event_name = 'purchase',       user_pseudo_id, NULL)) AS purchased
    FROM `bigquery-public-data.ga4_obfuscated_sample_ecommerce.events_*`
    WHERE _TABLE_SUFFIX BETWEEN '20201101' AND '20210131'
    GROUP BY medium
    ORDER BY users DESC
  "

  ga4_raw <- bq_project_query(billing_project, sql) |> bq_table_download()
  dir.create("data", showWarnings = FALSE)
  write_csv(ga4_raw, data_file)
}

ga4_raw <- read_csv(data_file, show_col_types = FALSE)
```

```{r}
#| label: wrangle
#| include: false
channel_lookup <- c(
  "organic"  = "Organic search",
  "cpc"      = "Paid search (CPC)",
  "referral" = "Referral",
  "(none)"   = "Direct"
)

ga4 <- ga4_raw |>
  filter(medium %in% names(channel_lookup)) |>
  mutate(
    channel   = unname(channel_lookup[medium]),
    conv_rate = purchased / users,
    share     = users / sum(users)
  ) |>
  arrange(desc(users))

funnel <- ga4 |>
  select(channel, users, viewed_item, added_to_cart, began_checkout, purchased) |>
  pivot_longer(-channel, names_to = "step", values_to = "n") |>
  mutate(step = factor(step,
    levels = c("users", "viewed_item", "added_to_cart", "began_checkout", "purchased"),
    labels = c("Visited", "Viewed item", "Added to cart", "Began checkout", "Purchased"))) |>
  group_by(channel) |>
  mutate(pct_of_visitors = n / max(n)) |>
  ungroup()

# Numbers used with inline code in the narrative
total_users      <- sum(ga4_raw$users)
total_purchasers <- sum(ga4_raw$purchased)
overall_cr       <- total_purchasers / total_users
cr_organic <- ga4$conv_rate[ga4$medium == "organic"]
cr_cpc     <- ga4$conv_rate[ga4$medium == "cpc"]
n_cpc      <- ga4$users[ga4$medium == "cpc"]
n_organic  <- ga4$users[ga4$medium == "organic"]

# Chi-square test: is purchasing independent of organic vs. paid channel?
chi_tab <- ga4 |>
  filter(medium %in% c("organic", "cpc")) |>
  transmute(channel, purchased, not_purchased = users - purchased)
chi_mat <- as.matrix(chi_tab[, -1])
rownames(chi_mat) <- chi_tab$channel
chi_res <- chisq.test(chi_mat)
```

# Introduction

## Who Is the Client? {transition="fade"}

:::::::: columns
:::: {.column width="58%"}
The **CPP Farm Store at Kellogg Ranch** is a Cal Poly Pomona operation that sells:

::: incremental
- Fresh produce and eggs grown on campus
- Plants and specialty products
- Customizable gift baskets
- Products both **in store** and through an **online store** linked from cpp.com
:::
::::

::::: {.column width="42%"}
:::: {.fragment .fade-left}
::: takeaway
A university-run store that has to balance an **educational mission** with the practical need to **generate revenue**.
:::
::::
:::::
::::::::

::: notes
Open by grounding the audience in who the client is. Emphasize that the online store is new, which is why search visibility matters now.
:::

## Problem Statement

The Farm Store's online store and cpp.com page currently have **minimal search visibility**.

. . .

It is unclear whether **paid search** (Google Shopping Ads) would meaningfully outperform **organic optimization** (SEO) alone.

. . .

Without conversion data, any ad budget decision would be a [guess]{.fragment .highlight-red}.

::: aside
Problem surfaced in the client kickoff meeting, Fall 2026. GA4 access was approved by the client on Sept. 29, 2026.
:::

::: notes
Pause after each line. The highlighted word "guess" is the hook for why analytics is needed.
:::

## Project and Analytics Objective {background-color="#2E5E3E"}

:::: {style="color: white;"}
**PO3:** Optimize the websites (cpp.com and the online store) by planning SEO strategies and implementing them.

::: {.fragment .grow}
**AO4:** Analyze GA4 conversion path data to evaluate whether Google Shopping Ads would improve traffic-to-purchase conversion.
:::
::::

::: notes
This is the single analytics objective this deck focuses on. The other three AOs belong to my teammates.
:::

## Why This Matters

::: incremental
- **Budget:** a small university store cannot afford to spend on ads that do not convert
- **Evidence:** the client asked for more data-driven decision making
- **Timing:** the online store is new, so a baseline set now makes every future campaign measurable
- **Decision:** results tell the Farm Store whether to invest in paid Shopping Ads, strengthen organic SEO, or do both
:::

# Literature Review

## Well-Established Models {.scrollable}

::: panel-tabset
### Paid search works

Keyword-targeted sponsored search can raise both click-through and conversion rates for online retailers, especially when ad placement and keyword relevance are optimized together [@ghose2009].

Shopping goal theory explains why advertisers bid to intercept customers at different stages of the shopping journey [@azemiozuem2026].

### What drives conversion

Not every campaign setting matters equally. **Budget** and **keyword matching** have the largest effect on conversions, while click-through rate and bid amount matter less [@olbrichschultz2014].

### Trust and value

For Google Shopping Ads specifically, purchase intention depends less on ad exposure and more on the **trust** and **perceived value** consumers attach to the ad, and this differs between Gen Y and Gen Z [@duffett2025].

### Measurement

GA4 reports the metrics this AO needs (traffic source, conversion rate, organic vs. paid acquisition), but it shows *what* happens in the funnel, not *why* [@mohaimin2025].
:::

::: notes
Click through each tab. The story goes: paid search can work, here is what makes it work, here is the consumer-side caveat, and here is the tool we will use to measure it.
:::

## Is the Data Adequate?

::::: columns
::: {.column width="50%"}
### What GA4 can answer

- Click-through and conversion rate
- Cost per click and ROAS (via Google Ads link)
- Organic vs. paid traffic split
- Where visitors drop off in the funnel
:::

::: {.column width="50%"}
### What GA4 cannot answer

- Do shoppers **trust** the ads?
- Do they see **value** in them?
- Which **keyword match types** drive results?
:::
:::::

:::: fragment
::: takeaway
**Need for additional analysis:** a short customer survey on ad trust and perceived value, plus a review of keyword-matching data once Google Ads is connected.[^1]
:::
::::

[^1]: Without these, conclusions would rest on click and conversion metrics alone, which prior research suggests is an incomplete picture of purchase intention.

# Methods

## Data and Sampling

:::::: columns
::: {.column width="55%"}
- **Population:** visitors to the Farm Store website and online store
- **Planned data:** Farm Store GA4 property (access pending)
- **Demo data for this proposal:** Google's public GA4 sample e-commerce dataset (Google Merchandise Store), Nov 2020 to Jan 2021
- **Sample:** all users in that period, **`r comma(total_users)` users**, of whom **`r comma(total_purchasers)`** made a purchase
:::

:::: {.column width="45%"}
::: callout-note
## Why a sample dataset?

The Farm Store GA4 property is not connected yet. The public dataset has the **exact same GA4 schema**, so the code written now will run on Farm Store data later by changing one table name.
:::
::::
::::::

::: notes
The sample-size numbers on this slide are generated by inline R code, so they update automatically if the data changes.
:::

## Data Wrangling {#data-wrangling .scrollable}

::: panel-tabset
### 1. Import

GA4 stores one row per **event**. A BigQuery SQL query rolls events up to one count per traffic medium and funnel step, and the result is cached as a CSV for reproducibility. [See full query](#sec-app-sql){.appendix-link}

```{r}
#| label: show-import
#| eval: false
ga4_raw <- bq_project_query(billing_project, sql) |> bq_table_download()
write_csv(ga4_raw, "data/ga4_channel_funnel.csv")
```

### 2. Clean

```{r}
#| label: show-clean
#| eval: false
ga4 <- ga4_raw |>
  filter(medium %in% names(channel_lookup)) |>     # drop "(data deleted)" and "<Other>"
  mutate(channel   = unname(channel_lookup[medium]),
         conv_rate = purchased / users)
```

### 3. Reshape

```{r}
#| label: show-reshape
#| eval: false
funnel <- ga4 |>
  select(channel, users:purchased) |>
  pivot_longer(-channel, names_to = "step", values_to = "n")
```
:::

## Sample Characteristics

```{r}
#| label: tbl-sample
#| echo: false
#| tbl-cap: "Users and purchasers by traffic channel (GA4 sample data)"
ga4 |>
  transmute(
    Channel           = channel,
    Users             = comma(users),
    `Share of users`  = percent(share, 0.1),
    Purchasers        = comma(purchased),
    `Conversion rate` = percent(conv_rate, 0.01)
  ) |>
  kable(align = "lrrrr")
```

::: fragment
@tbl-sample shows paid search brought in `r comma(n_cpc)` users compared with `r comma(n_organic)` from organic search.
:::

## Conversion by Channel {.smaller}

```{r}
#| label: fig-conv
#| fig-cap: "Visitor-to-purchase conversion rate by channel"
#| output-location: column
#| fig-width: 7
#| fig-height: 5
ga4 |>
  ggplot(aes(x = reorder(channel, conv_rate),
             y = conv_rate)) +
  geom_col(fill = "#2E5E3E") +
  geom_text(aes(label = percent(conv_rate, 0.01)),
            hjust = -0.1, size = 6) +
  coord_flip() +
  scale_y_continuous(labels = percent,
                     expand = expansion(mult = c(0, 0.2))) +
  labs(x = NULL,
       y = "Conversion rate") +
  theme_minimal(base_size = 18)
```

::: notes
Code on the left, chart on the right. This is the reproducible chart required for the assignment. Point out that the chart rebuilds itself every time the deck is rendered.
:::

## Where Visitors Drop Off

```{r}
#| label: fig-funnel
#| echo: false
#| fig-cap: "Share of each channel's visitors who reach each funnel step"
#| fig-width: 13
#| fig-height: 5.5
funnel |>
  ggplot(aes(step, pct_of_visitors, color = channel, group = channel)) +
  geom_line(linewidth = 1.3) +
  geom_point(size = 3) +
  scale_y_continuous(labels = percent) +
  scale_color_manual(values = c("#2E5E3E", "#FFB81C", "#6B4E31", "#8FA98F")) +
  labs(x = NULL, y = "% of channel visitors", color = NULL) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "top")
```

::: notes
@fig-funnel is the "conversion path" view. Look for the step where one line separates from the others.
:::

## Measures

| Construct | GA4 / Google Ads metric | Scale |
|--------------------------|---------------------------|-------------------|
| Traffic acquisition | Users by medium (organic, cpc, referral, direct) | Nominal |
| Engagement | View item, add to cart, begin checkout | Ratio (counts) |
| Conversion | Purchasers / users | Ratio (%) |
| Ad efficiency | Click-through rate, cost per click | Ratio |
| Ad return | Return on ad spend (ROAS) | Ratio |
| Ad trust and perceived value | Survey items (planned) | Interval (Likert) |

: Constructs and how they are operationalized

## Analytics Methods

::: incremental
1.  **Descriptive funnel analysis** by channel (@fig-funnel)
2.  **Chi-square test of independence:** is purchasing independent of organic vs. paid channel?
3.  **Pre/post comparison** once the Farm Store runs a Shopping Ads test period
4.  **Survey analysis** of ad trust and perceived value as drivers of purchase intention
:::

. . .

::: takeaway
**Preview on sample data:** paid search converted at `r percent(cr_cpc, 0.01)` vs. `r percent(cr_organic, 0.01)` for organic search, $\chi^2$ = `r round(chi_res$statistic, 2)`, p `r ifelse(chi_res$p.value < .001, "< .001", paste("=", round(chi_res$p.value, 3)))`.
:::

::: fragment
In Google's store, paid traffic converted `r ifelse(cr_cpc < cr_organic, "*lower*", "*higher*")` than organic search. `r ifelse(cr_cpc < cr_organic, "This fits the trust and perceived value argument: ad clicks alone do not guarantee purchases [@duffett2025]. It is exactly the question the Farm Store data needs to answer before any ad budget is spent.", "This suggests paid search can pull high-intent shoppers, consistent with prior sponsored search research [@ghose2009]. The Farm Store data will show whether the same holds for a small local store.")`
:::

::: notes
Stress that these numbers come from Google's demo store, not the Farm Store. They show the method works; the Farm Store results will come once GA4 is connected.
:::

# Wrap Up

## Next Steps {transition="zoom"}

::: {.fragment .fade-up}
1.  Connect the Farm Store GA4 property (client approved, IT setup pending)
:::

::: {.fragment .fade-up}
2.  Swap the sample table name for the Farm Store GA4 export and rerun
:::

::: {.fragment .fade-up}
3.  Draft the ad trust and perceived value survey items
:::

::: {.fragment .fade-up}
4.  Recommend: paid Shopping Ads, stronger organic SEO, or both
:::

## References

::: {#refs}
:::

# Appendix {visibility="uncounted"}

## A. Full BigQuery SQL {#sec-app-sql visibility="uncounted"}

```{r}
#| label: show-sql
#| echo: false
#| comment: ""
cat(
"SELECT
  traffic_source.medium AS medium,
  COUNT(DISTINCT user_pseudo_id) AS users,
  COUNT(DISTINCT IF(event_name = 'view_item',      user_pseudo_id, NULL)) AS viewed_item,
  COUNT(DISTINCT IF(event_name = 'add_to_cart',    user_pseudo_id, NULL)) AS added_to_cart,
  COUNT(DISTINCT IF(event_name = 'begin_checkout', user_pseudo_id, NULL)) AS began_checkout,
  COUNT(DISTINCT IF(event_name = 'purchase',       user_pseudo_id, NULL)) AS purchased
FROM `bigquery-public-data.ga4_obfuscated_sample_ecommerce.events_*`
WHERE _TABLE_SUFFIX BETWEEN '20201101' AND '20210131'
GROUP BY medium
ORDER BY users DESC")
```

[Back to Data Wrangling](#data-wrangling){.appendix-link}

## B. Project Files {visibility="uncounted"}

- [Live presentation (GitHub Pages)](https://jukedz.github.io/ibm6500-assignments/M06-1-Revealjs-Presentation/farm-store-ao4-presentation.html)
- [Source code and data (GitHub repo folder)](https://github.com/jukedz/ibm6500-assignments/tree/main/M06-1-Revealjs-Presentation)
- [Quarto source file (.qmd)](https://github.com/jukedz/ibm6500-assignments/blob/main/M06-1-Revealjs-Presentation/farm-store-ao4-presentation.qmd)
- [Cached GA4 query result (.csv)](https://github.com/jukedz/ibm6500-assignments/blob/main/M06-1-Revealjs-Presentation/data/ga4_channel_funnel.csv)
- [Custom theme (.scss)](https://github.com/jukedz/ibm6500-assignments/blob/main/M06-1-Revealjs-Presentation/farmstore.scss)

## C. Useful Links {visibility="uncounted"}

- [Google's GA4 sample e-commerce dataset documentation](https://developers.google.com/analytics/bigquery/web-ecommerce-demo-dataset)
- [GA4 BigQuery export schema](https://support.google.com/analytics/answer/7029846)
- [CPP Farm Store](https://www.cpp.edu/farmstore/)
- [Quarto Revealjs documentation](https://quarto.org/docs/presentations/revealjs/)
- [CCIDM effective Quarto slides repo](https://github.com/CCIDM/effective-quarto-slides)